\propertitle

### 2019-11-26 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-26T14:07:21+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

In our case we don't know the time averages of the quantity
$\binom{\yA}{m}$, so we cannot directly equate them to the factorial moment
$\E\bigl[\binom{\yA}{m}\bigr]$ of the larger-population distribution
$\pxxx(\yA)$. But \eqn~\eqref{eq:equal_m_factorialmoments} of
step~\ref{item:sample_step} comes to our rescue, because it says that the
expectation for the larger population are determined by that for the sample
$\E\bigl[\binom{\ya}{m}\bigr]$, and we do have the time average of its
corresponding sample quantity $\binom{\ya}{m}$. So we can combine the two
steps:
\begin{equation*}
  \label{eq:chain_prescription_moments}
  \mathrlap{\overbracket{\phantom{\text{measured moments} \rightarrow
        \text{sample moments}}}^{\textit{maximum-entropy prescription}}}
  \text{measured moments} \rightarrow
        \underbracket{\text{sample moments}
  \rightarrow
    \text{larger-population moments}}_{\textit{sampling theory}}
\end{equation*}
We obtain a distribution $\px(\yA \| M, N)$ for the larger population of
$N$ neurons by constraining some of its factorial moments, for example
$M=\set{1,\dotsc,m'}$ with $m'\le n$, to be equal to the sample's recorded
averages. In formulae, the constraints on $\pxxx(\yA)$ are
\begin{equation}
  \label{eq:final_constraint}
\underbracket{\frac{1}{T}\sum_{t}\binom{\ya_{t}}{m}\bigg/\binom{n}{m}
  \equiv
   \sum_{\ya}
  \binom{\ya}{m}\bigg/  \binom{n}{m} \; \yff_{\ya}}_{\text{measured moments}}
  \mathrel{\;=\;}
   \underbracket{\sum_{\yA}
     \binom{\yA}{m}\bigg/  \binom{N}{m} \; \pxxx(\yA)}_{\text{distribution moments}},
   \quad m\in M.
\end{equation}

@@

The amount and degrees of the constraining moments depend on the questions
and hypotheses that a researcher is exploring. We give some examples in the
next two sections. Note that the constraint
equation~\eqref{eq:final_constraint} for the
distribution~\eqref{eq:app_maxent_pop_first} may not have exact solutions
if $N$ is strictly larger than $n$; and the possible discrepancy typically
increases with the number of constraints. This discrepancy comes from the
approximate character of two assumptions behind the maximum-entropy method
and of one assumption specific to our application: that the order of the
time bins is irrelevant, that measured moments equal expectations (or
equivalently, the number of time bins $T$ is infinite), and that an
activity level $\yA$ can equally likely be generated by any set of $\yA$
neurons in the larger population. The magnitude of this discrepancy can be
a signature of the (at least temporary) presence of a \enquote{neuronal
  assembly} \citep[\chap~12]{gerstneretal2014}{hebb1949_r2002}. We discuss
this matter in \sect~\ref{sec:assumptions}.

@@

\enquote{Cooperativity}, \enquote{interaction}, and similar terms are
vague, so we need to translate them into a more precise notion first. Here
we use the notion of \emph{informational sufficiency}
\citep[\sect~4.5]{bernardoetal1994}[\chap~8 \amp\
\sect~14.2]{jaynes1994_r2003}{cifarellietal1982,kullbacketal1951}[the
notion goes back to][]{fisher1922} because it relates to those terms, is
intuitive, and is connected with maximum-entropy distributions. Its idea is
as follows. Our probabilities about the frequencies of the activities of
the sample, or about the activity of the sample in a new time bin, are in
principle conditional on all experimental data and statistics we have. But
it can be the case that discarding part of the data or statistics -- for
example, the measured third- and higher-order moments -- leaves our
probabilities almost unchanged. This means that the discarded statistics
are \emph{informationally irrelevant} or almost so. The remaining
statistics -- for example, first and second moments -- are
\emph{informationally sufficient}.\footnote{For more technical results
  and connections with the notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,lauritzen1982_r1988,lauritzen2007,cifarellietal1980,cifarellietal1981,diaconisetal1981,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,ayetal2015}.}
% This informational quasi-sufficiency is interesting because
% it may give clues about population organization and dynamics.

@@

Before applying this notion to neuronal activity, however, we must keep in
mind that \emph{informational sufficiency is not preserved under sampling}:
if a probability distribution has some sufficient statistics, then its
marginals, such as the distribution for a sample, \emph{cannot} have the
same sufficient statistics, and vice versa; except for trivial cases such
as uniform probability distributions. % This generally holds also among
% different marginals: if a statistics is sufficient for some marginal, then
% it will not be sufficient for another marginal.
This impossibility is known in statistical mechanics: if a system is
described by a Gibbs state, its subsystems cannot be perfectly described by
Gibbs states \citep[\eg][and references therein]{maesetal1999}.
Mathematically this impossibility comes from the Pitman-Koopman theorem
mentioned above, and translates into the general impossibility of solving a
system of $n$ independent equations in $m$ unknowns with more equations
than unknowns, $n>m$ \citep[\sect~3.1]{portamanaetal2015}.

This fact is important for our analysis. If, say, means and pairwise
correlations seem informationally sufficient for a particular sample from a
brain area, then they may well not be sufficient for the larger population of
neurons constituting that area, and vice versa. So if what interests us is
\enquote{cooperativity} or \enquote{interaction} of a brain area, it is
unreliable to use a maximum-entropy distribution constructed only for the
sample. % \citep[contrast this with some literature, \eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}
The approach presented here avoids this problem, because the
maximum-entropy method is applied to obtain the distribution of the larger
population, not of the sample alone.

@@

We measure the difference $\ydi(M'',M')$ in informational sufficiency
between a set of moments, say $M''\defd\set{1,\dotsc,m''}$, and another, say
$M'\defd \set{1,\dotsc,m'}$, as follows:
\begin{enumerate}[label=(\roman*)]
\item from each maximum-entropy distributions $\px(\yA \| M)$ for the larger
  population, built from each set of constraints $M=M',M''$, calculate the
  marginal distribution for the sample:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya \| M) = \sum_{\yA}\yG_{\ya \yA}\px(\yA \| M),
  \qquad M=M',M'';
\end{equation}
\item calculate the relative entropies of the measured frequency
  distribution $\yf$ with respect to each sample marginal, and multiply
  them by the number of time bins $T$:
  \begin{equation}
    \label{eq:relentropy_T}
    T\,\sh[\yf; \pxx(\ya \| M)] \defd  T  \sum_{\ya} \yff_{\ya} \log\frac{\yff_{\ya}}{
      \pxx(\ya \| M)},
    \qquad M=M',M'';    
  \end{equation}
\item take the difference: 
\begin{multline}
  \label{eq:measure_suffic}
  \ydi(M'',M') \defd
  T \,\sh[\yf; \pxx(\ya \| M')]
  -
  T \,\sh[\yf; \pxx(\ya \| M'')]
  \equiv{}\\
T  \sum_{\ya} \yff_{\ya} \log\frac{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M'')}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M')}.
\end{multline}
\end{enumerate}
The measure $\ydi(M'',M')$ so defined is positive if $M''$ is \enquote{more
  informationally sufficient} than $M'$, and negative otherwise.

Why is this a natural measure? Because $\ydi(M'',M')$ is equal to the
log-ratio of the probabilities of the data $\yf$ conditional on the
hypotheses $M''$ and $M'$:
\begin{equation}
  \label{eq:bayes_factor}
  \ydi(M'',M') =
  \log\bigl[\pf(\yf \| M'')/
  \pf(\yf \| M')\bigr].
\end{equation}
This is called their \emph{relative weight of evidence}, the logarithm of
their \emph{relative Bayes factor}
\citep[\chap~6]{good1950}{good1975,good1981,good1985,good1983}[\sect~1.4]{osteyeeetal1974}{mackay1992,kassetal1995}[see
also][p.~421]{jeffreys1936}[\chaps~V, VI, A]{jeffreys1939_r1983}. We prove
this equality in appendix~\ref{sec:measure_suff}. So the exponential of
$\ydi(M'',M')$ tells us how much more probable the data $\yf$ are
conditional on $M''$, than conditional on $M'$. We can also combine this
measure with pre-data probabilities for the two hypotheses to obtain the
ratio of their probabilities conditional on the data
\citep[\cf][]{bretthorst2013}.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-11-26T14:21:20+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data {\large\color{myred}[draft]}%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

In our case we don't know the time averages of the quantity
$\binom{\yA}{m}$, so we cannot directly equate them to the factorial moment
$\E\bigl[\binom{\yA}{m}\bigr]$ of the larger-population distribution
$\pxxx(\yA)$. But \eqn~\eqref{eq:equal_m_factorialmoments} of
step~\ref{item:sample_step} comes to our rescue, because it says that the
expectation for the larger population are determined by that for the sample
$\E\bigl[\binom{\ya}{m}\bigr]$, and we do have the time average of its
corresponding sample quantity $\binom{\ya}{m}$. So we can combine the two
steps:
\begin{equation*}
  \label{eq:chain_prescription_moments}
  \mathrlap{\overbracket{\phantom{\text{measured averages} \rightarrow
        \text{sample moments}}}^{\textit{maximum-entropy prescription}}}
  \text{measured moments} \rightarrow
        \underbracket{\text{sample moments}
  \rightarrow
    \text{larger-population moments}}_{\textit{sampling theory}}
\end{equation*}
We obtain a distribution $\px(\yA \| M, N)$ for the larger population of
$N$ neurons by constraining some of its factorial moments, for example
$M=\set{1,\dotsc,m'}$ with $m'\le n$, to be equal to the sample's recorded
averages. In formulae, the constraints on $\pxxx(\yA)$ are
\begin{equation}
  \label{eq:final_constraint}
\underbracket{\frac{1}{T}\sum_{t}\binom{\ya_{t}}{m}\bigg/\binom{n}{m}
  \equiv
   \sum_{\ya}
  \binom{\ya}{m}\bigg/  \binom{n}{m} \; \yff_{\ya}}_{\text{measured averages}}
  \mathrel{\;=\;}
   \underbracket{\sum_{\yA}
     \binom{\yA}{m}\bigg/  \binom{N}{m} \; \pxxx(\yA)}_{\text{distribution moments}},
   \quad m\in M.
\end{equation}

@@

The amount and degrees of the constraining moments depend on the questions
and hypotheses that a researcher is exploring. We give some examples in the
next two sections. Note that the constraint
equation~\eqref{eq:final_constraint} for the
distribution~\eqref{eq:app_maxent_pop_first} may not have exact solutions
if $N$ is strictly larger than $n$; and the possible discrepancy typically
increases with the number of constraints. This discrepancy comes from the
approximate character of two assumptions behind the maximum-entropy method
and of one assumption specific to our application: that the order of the
time bins is irrelevant, that measured averages equal expected values (or
equivalently, the number of time bins $T$ is infinite), and that an
activity level $\yA$ can equally likely be generated by any set of $\yA$
neurons in the larger population. The magnitude of this discrepancy can be
a signature of the (at least temporary) presence of a \enquote{neuronal
  assembly} \citep[\chap~12]{gerstneretal2014}{hebb1949_r2002}. We discuss
this matter in \sect~\ref{sec:assumptions}.

@@

\enquote{Cooperativity}, \enquote{interaction}, and similar terms are
vague, so we need to translate them into a more precise notion first. Here
we use the notion of \emph{informational sufficiency}
\citep[\sect~4.5]{bernardoetal1994}[\chap~8 \amp\
\sect~14.2]{jaynes1994_r2003}{cifarellietal1982,kullbacketal1951}[the
notion goes back to][]{fisher1922} because it relates to those terms, is
intuitive, and is connected with maximum-entropy distributions. Its idea is
as follows. Our probabilities about the frequencies of the activities of
the sample, or about the activity of the sample in a new time bin, are in
principle conditional on all experimental data and statistics we have. But
it can be the case that discarding part of the data or statistics -- for
example, the third- and higher-order empirical moments -- leaves our
probabilities almost unchanged. This means that the discarded statistics
are \emph{informationally irrelevant} or almost so. The remaining
statistics -- for example, first and second empirical moments -- are
\emph{informationally sufficient}.\footnote{For more technical results and
  connections with the notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,lauritzen1982_r1988,lauritzen2007,cifarellietal1980,cifarellietal1981,diaconisetal1981,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,ayetal2015}.}
% This informational quasi-sufficiency is interesting because
% it may give clues about population organization and dynamics.

@@

Before applying this notion to neuronal activity, however, we must keep in
mind that \emph{informational sufficiency is not preserved under sampling}.
If a probability distribution has some sufficient statistics, then its
marginals, such as the distribution for a sample, \emph{cannot} have the
same sufficient statistics, and vice versa; except for trivial cases such
as uniform probability distributions. % This generally holds also among
% different marginals: if a statistics is sufficient for some marginal, then
% it will not be sufficient for another marginal.
This impossibility is known in statistical mechanics: if a system is
described by a Gibbs state, its subsystems cannot be perfectly described by
Gibbs states \citep[\eg][and references therein]{maesetal1999}.
Mathematically this impossibility comes from the Pitman-Koopman theorem
mentioned above, and translates into the general impossibility of solving a
system of independent equations with more equations than unknowns
\citep[\sect~3.1]{portamanaetal2015}.

This fact is important for our analysis. If, say, means and pairwise
correlations seem informationally sufficient for a particular sample from a
brain area, then they may well not be sufficient for the larger population
of neurons constituting that area, and vice versa. So if we are interested
in the \enquote{cooperativity} or \enquote{interaction} of a brain area, it
is unreliable to use a maximum-entropy distribution constructed only for a
sample
thereof. % \citep[contrast this with some literature, \eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}
The approach presented here avoids this problem because the maximum-entropy
method is applied to obtain the distribution of the larger population, not
of the sample alone.

@@

We measure the difference $\ydi(M'',M')$ in informational sufficiency
between a set of moments, say $M''\defd\set{1,\dotsc,m''}$, and another, say
$M'\defd \set{1,\dotsc,m'}$, as follows:
\begin{enumerate}[label=(\roman*)]
\item from each maximum-entropy distributions $\px(\yA \| M,N)$ for the larger
  population, built from each set of constraints $M=M',M''$, calculate the
  marginal distribution for the sample:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya \| M, N) = \sum_{\yA}\yG_{\ya \yA}\px(\yA \| M,N),
  \qquad M=M',M'';
\end{equation}
\item calculate the relative entropies of the measured frequency
  distribution $\yf$ with respect to each sample marginal, and multiply
  them by the number of time bins $T$:
  \begin{equation}
    \label{eq:relentropy_T}
    T\,\sh[\yf; \pxx(\ya \| M,N)] \defd
    T  \sum_{\ya} \yff_{\ya} \log\frac{\yff_{\ya}}{\pxx(\ya \| M,N)},
    \qquad M=M',M'';    
  \end{equation}
\item take the difference: 
\begin{equation}
  \label{eq:measure_suffic}
  \begin{split}
  \ydi_{N}(M'',M') &\defd
  T \,\sh[\yf; \pxx(\ya \| M',N)]
  -
  T \,\sh[\yf; \pxx(\ya \| M'',N)]
  \\ &\equiv
T  \sum_{\ya} \yff_{\ya} \log\frac{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M'',N)}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M',N)}.
\end{split}
\end{equation}
\end{enumerate}
The measure $\ydi_{N}(M'',M')$ so defined is positive if $M''$ is \enquote{more
  informationally sufficient} than $M'$, and negative otherwise.

Why is this a natural measure? Because $\ydi_{N}(M'',M')$ is equal to the
log-ratio of the probabilities of the data $\yf$ conditional on the
hypotheses $M''$ and $M'$:
\begin{equation}
  \label{eq:bayes_factor}
  \ydi(M'',M') =
  \log\bigl[\pf(\yf \| M'')/
  \pf(\yf \| M')\bigr].
\end{equation}
This is called their \emph{relative weight of evidence}, the logarithm of
their \emph{relative Bayes factor}
\citep[\chap~6]{good1950}{good1975,good1981,good1985,good1983}[\sect~1.4]{osteyeeetal1974}{mackay1992,kassetal1995}[see
also][p.~421]{jeffreys1936}[\chaps~V, VI, A]{jeffreys1939_r1983}. We prove
this equality in appendix~\ref{sec:measure_suff}. So the exponential of
$\ydi(M'',M')$ tells us how much more probable the data $\yf$ are
conditional on $M''$, than conditional on $M'$. We can also combine this
measure with pre-data probabilities for the two hypotheses to obtain the
ratio of their probabilities conditional on the data
\citep[\cf][]{bretthorst2013}.
